---
output: beamer_presentation
header-includes: \usepackage{graphicx}
---


```{r setup, include = F}

rm(list=ls())

#Load knitr 
library(knitr)

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = '..')
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.width=8,
               fig.height=8,
               fig.path = "../Figures/")

```

```{r load and clean, include = F}

#Load and clean data
# source('Scripts/GWAS_sims_LOAD.R')
# source('Scripts/GWAS_sims_FUNCTIONS.R')
# source('Scripts/GWAS_sims_CLEAN.R')

```



## Genome-wide association studies in natural populations: managing expectations and avoiding error

Susan E. Johnston & **Lewis G. Spurgin**

\vspace{0.3cm}

\begin{center}
\includegraphics[width=0.8\textwidth]{Images/manhattan.png}
\end{center}

\vspace{0.3cm}

\begin{center}
\includegraphics[width=0.9\textwidth]{Images/logos.png}
\end{center}






##**G**enome **W**ide **A**ssociation **S**tudy (GWAS)

Experimental design used to detect associations between genetic variants and traits in population samples







----

\begin{center}
\includegraphics[width=1\textwidth]{Images/gwas_success.png}
\end{center}

\begin{flushright}
\tiny{Visscher et al. 2017, AJHG}
\end{flushright}





----

\begin{center}
\includegraphics[width=1\textwidth]{Images/traits.png}
\end{center}







##GWAS in natural populations

Example studies (finches, peppered moths etc)






##Key factors to consider in GWAS

- Architecture of the trait
- Number of individuals sampled
- Number of genetic markers sampled
- Extent of linkage disequilibrium between sampled markers and causal variants






##Key factors to consider in GWAS

- Architecture of the trait
- Number of individuals sampled
- Number of genetic markers sampled
- **Extent of linkage disequilibrium between sampled markers and causal variants**






##Linkage disequilibrium (LD) and GWAS

Why LD is important for GWAS




##Linkage disequilibrium (LD) and GWAS

Factors affecting LD




##Linkage disequilibrium (LD) and GWAS

LD in natural populations
